---
title: "Power Simulation: H1 — Vibe Coding × Time Interaction"
subtitle: "Longitudinal Assessment of AI-Assisted Coding and Python Skill"
author: "Andy Woods, Royal Holloway, University of London"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: kate
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  cache   = TRUE,        # cache slow simulation chunks
  fig.width  = 9,
  fig.height = 6,
  dpi     = 150
)

library(lme4)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(MASS)
library(knitr)
library(kableExtra)
library(parallel)    # base R — mclapply for multi-core parallelisation

# pbmcapply adds a progress bar to mclapply; install if missing
if (!requireNamespace("pbmcapply", quietly = TRUE)) install.packages("pbmcapply")
library(pbmcapply)

# MASS::select masks dplyr::select; prefer dplyr throughout.
if (!requireNamespace("conflicted", quietly = TRUE)) install.packages("conflicted")
library(conflicted)
# Pre-declare all known conflicts from lme4/Matrix/brms dependencies
# to avoid whack-a-mole errors during knitting.
conflicts_prefer(
  dplyr::select,
  dplyr::filter,
  dplyr::lag,
  tidyr::expand,
  tidyr::pack,
  tidyr::unpack,
  stats::ar,
  stats::sigma
)

# Format seconds as "Xm Ys" or "Xs"
format_time <- function(seconds) {
  seconds <- round(seconds)
  if (seconds < 60) return(sprintf("%ds", seconds))
  sprintf("%dm %02ds", seconds %/% 60L, seconds %% 60L)
}
```

---

## Background

This document powers the primary hypothesis of the vibe coding study (see `research_plan.md §8`):

> **H1:** Participants who report a higher vibe-coding percentage will show a steeper decline in challenge accuracy over time.

The key quantity is the interaction term `vibe_coding_pct × months_since_baseline` in a multilevel model with accuracy (logit-transformed) as the outcome.

### Minimum Effect Size of Interest (MESI)

From `research_plan.md §8.2`:

- A **≥ 5 percentage-point** accuracy decline per 6 months at `vibe_coding_pct = 1.0` vs `0.0`, all else equal.
- 6 months = 2 time units (1 unit = 3 months).
- On the logit scale: $\text{logit}(0.65) - \text{logit}(0.70) \approx -0.228$ across 2 units at full vibe coding.
- Interaction coefficient: $-0.228 / 2 = -0.114$ per 3-month unit per 100% vibe coding.

---

## 1. Data-Generating Parameters

```{r params}
# Fixed effects (logit scale)
INTERCEPT        <- 0.847   # logit(0.70) — ~70% baseline accuracy
BETA_TIME        <- 0.05    # slight improvement from practice alone, per 3-month unit
BETA_VIBE        <- -0.05   # direct effect of vibe coding within a session
BETA_DIFFICULTY  <- -0.30   # per difficulty tier (1–5)

# KEY INTERACTION (MESI)
BETA_INTERACTION <- -0.114  # per 3-month unit per 100% vibe coding

# Random effects
SD_INTERCEPT_P <- 0.50   # between-person SD in baseline ability
SD_SLOPE_P     <- 0.15   # between-person SD in skill trajectory
COR_P          <- -0.20  # faster improvers tend to start lower
SD_CHALLENGE   <- 0.40   # between-challenge variation beyond tier rating
```

The MESI implies the following accuracy trajectories for a participant at average ability:

```{r mesi-illustration, echo=FALSE}
tibble(
  months          = seq(0, 12, by = 1.5),
  time_unit       = months / 3,
  vibe_0pct       = plogis(INTERCEPT + BETA_TIME  * time_unit),
  vibe_50pct      = plogis(INTERCEPT + (BETA_TIME + BETA_INTERACTION * 0.5) * time_unit + BETA_VIBE * 0.5),
  vibe_100pct     = plogis(INTERCEPT + (BETA_TIME + BETA_INTERACTION * 1.0) * time_unit + BETA_VIBE * 1.0)
) |>
  pivot_longer(starts_with("vibe_"), names_to = "group", values_to = "accuracy") |>
  mutate(group = recode(group,
    vibe_0pct   = "0% vibe coding",
    vibe_50pct  = "50% vibe coding",
    vibe_100pct = "100% vibe coding"
  )) |>
  ggplot(aes(x = months, y = accuracy, colour = group)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0.4, 1)) +
  scale_colour_brewer(palette = "Set2", name = NULL) +
  labs(
    title    = "Simulated accuracy trajectories at the MESI",
    subtitle = "Tier 1 challenge, average-ability participant",
    x        = "Months since baseline",
    y        = "Expected accuracy"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

---

## 2. Data Simulation

The simulated dataset reflects the expected study structure:

- **Sessions per participant:** Poisson-distributed (minimum 2), as participants may only complete baseline + 1 follow-up.
- **Challenges per session:** 3–6 (weighted toward 4–5), reflecting expected dropout before the harder challenges.
- **Vibe-coding distribution:** bimodal — low adopters (0–25%) and heavy adopters (50–100%).
- **Difficulty tier:** ascending within session (positions 1–3 = Tier 1, 4–6 = Tier 2, 7–8 = Tier 3); Tier 3+ is rarely reached.

```{r simulate-fn}
# Fully vectorised — no nested for-loops. Generates all participants, sessions,
# and challenges in bulk using expand/rep, then computes eta in a single pass.
simulate_dataset <- function(n_participants,
                              mean_sessions      = 5.0,
                              session_gap_months = 1.5) {

  # --- Participant level ---
  re_cov <- matrix(
    c(SD_INTERCEPT_P^2,
      COR_P * SD_INTERCEPT_P * SD_SLOPE_P,
      COR_P * SD_INTERCEPT_P * SD_SLOPE_P,
      SD_SLOPE_P^2),
    nrow = 2
  )
  re_p <- MASS::mvrnorm(n_participants, mu = c(0, 0), Sigma = re_cov)

  n_sess    <- pmax(2L, rpois(n_participants, mean_sessions))
  vibe_base <- ifelse(
    runif(n_participants) < 0.5,
    runif(n_participants, 0.00, 0.25),   # low adopters
    runif(n_participants, 0.50, 1.00)    # heavy adopters
  )

  # --- Session level (expand participant → sessions) ---
  pid      <- rep(seq_len(n_participants), n_sess)
  sess_idx <- sequence(n_sess)                      # 1,2,...,n_sess[i] per participant

  months    <- (sess_idx - 1) * session_gap_months
  time_unit <- months / 3
  vibe_pct  <- pmin(1, pmax(0, vibe_base[pid] + rnorm(length(pid), 0, 0.05)))
  n_chal    <- sample(3:6, length(pid), replace = TRUE, prob = c(0.10, 0.30, 0.40, 0.20))

  # --- Challenge level (expand sessions → challenges) ---
  sess_row  <- rep(seq_along(pid), n_chal)
  c_idx     <- sequence(n_chal)

  tier <- ifelse(c_idx <= 3, 1L,
          ifelse(c_idx <= 6, 2L,
          ifelse(c_idx <= 8, 3L,
          ifelse(c_idx == 9, 4L, 5L))))

  # --- Vectorised linear predictor ---
  eta <- INTERCEPT +
    re_p[pid[sess_row], 1] +
    (BETA_TIME + re_p[pid[sess_row], 2]) * time_unit[sess_row] +
    BETA_VIBE       * vibe_pct[sess_row] +
    BETA_DIFFICULTY * (tier - 1) +
    BETA_INTERACTION * vibe_pct[sess_row] * time_unit[sess_row] +
    rnorm(length(sess_row), 0, SD_CHALLENGE)

  prob         <- plogis(eta)
  tests_passed <- rbinom(length(prob), 5L, prob)
  accuracy     <- tests_passed / 5

  tibble(
    participant_id        = factor(pid[sess_row]),
    session               = sess_idx[sess_row],
    challenge_idx         = c_idx,
    months_since_baseline = months[sess_row],
    time_unit             = time_unit[sess_row],
    vibe_coding_pct       = vibe_pct[sess_row],
    difficulty_tier       = tier,
    accuracy              = accuracy,
    accuracy_logit        = log((accuracy + 0.01) / (1 - accuracy + 0.01))
  ) |>
    mutate(
      vibe_coding_pct_c = vibe_coding_pct - mean(vibe_coding_pct),
      difficulty_tier_c = difficulty_tier  - mean(difficulty_tier)
    )
}
```

### Example dataset (N = 50)

```{r example-dataset}
set.seed(20260219)
example_dat <- simulate_dataset(n_participants = 50)

glimpse(example_dat)
```

```{r example-distributions, echo=FALSE, fig.height=4}
p1 <- example_dat |>
  distinct(participant_id, session, vibe_coding_pct) |>
  distinct(participant_id, .keep_all = TRUE) |>
  ggplot(aes(x = vibe_coding_pct)) +
  geom_histogram(bins = 20, fill = "#66c2a5", colour = "white") +
  labs(title = "Vibe-coding % distribution", x = "Vibe-coding %", y = "Participants") +
  theme_minimal()

p2 <- example_dat |>
  distinct(participant_id, session) |>
  count(participant_id) |>
  ggplot(aes(x = n)) +
  geom_bar(fill = "#fc8d62", colour = "white") +
  scale_x_continuous(breaks = 1:10) +
  labs(title = "Sessions per participant", x = "N sessions", y = "Participants") +
  theme_minimal()

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

---

## 3. Model and Test

For speed, each simulated dataset is fitted with `lme4` (Wald t-test). This approximates the `brms` Bayesian model specified in `research_plan.md §5`. Power is the proportion of simulations where $|t| > 1.96$ (two-tailed $\alpha = .05$).

```{r fit-fn}
fit_and_test <- function(dat) {
  # Use || (zero-correlation parameterisation) to decouple random intercept and
  # slope — one fewer parameter, greatly reduces singular fits when participants
  # have only 2 sessions. The fixed-effect interaction t-value (our test
  # statistic) is unaffected by this simplification.
  # suppressWarnings: boundary/singular warnings are expected in simulation
  # contexts with small per-person sample sizes and are not informative here.
  m <- tryCatch(
    suppressWarnings(
      lmer(
        accuracy_logit ~ vibe_coding_pct_c * time_unit +
          difficulty_tier_c +
          (1 + time_unit || participant_id),
        data    = dat,
        control = lmerControl(optimizer = "bobyqa")
      )
    ),
    error = function(e) NULL
  )

  if (is.null(m)) return(NA_real_)

  coefs <- summary(m)$coefficients
  row   <- grep("vibe_coding_pct_c:time_unit", rownames(coefs))
  if (length(row) == 0) return(NA_real_)

  coefs[row, "t value"]
}
```

---

## 4. Power Sweep

```{r power-fn}
run_power_sim <- function(n_participants,
                           mean_sessions = 2.5,
                           n_sims        = 500,
                           alpha         = 0.05) {
  t_crit <- qnorm(1 - alpha / 2)

  t_vals <- replicate(n_sims, {
    dat <- simulate_dataset(n_participants, mean_sessions = mean_sessions)
    fit_and_test(dat)
  })

  t_vals <- t_vals[!is.na(t_vals)]
  power  <- mean(abs(t_vals) > t_crit)
  se     <- sqrt(power * (1 - power) / length(t_vals))

  tibble(
    n_participants = n_participants,
    mean_sessions  = mean_sessions,
    n_sims_valid   = length(t_vals),
    power          = power,
    power_lower    = max(0, power - 1.96 * se),
    power_upper    = min(1, power + 1.96 * se)
  )
}
```

**Expected runtime:** ~2–4 minutes on an M4 Max (25 conditions × 500 simulations, parallelised across 15 cores; vectorised data generation eliminates the nested for-loop bottleneck). The chunk is cached, so re-knitting is instant. A progress bar appears in the RStudio console/terminal during the run; it is not captured in the knitted HTML.

```{r run-sweep, cache=TRUE}
# L'Ecuyer-CMRG gives each forked process its own independent RNG stream,
# ensuring reproducibility under parallel execution.
RNGkind("L'Ecuyer-CMRG")
set.seed(20260219)

# Mean sessions reflect expected distribution: minimum 2, most completing ~6.
# Poisson(mean) with max(2, .) gives realistic unbalanced data.
# Range covers pessimistic dropout (3) through engaged participants (8).
param_grid <- expand.grid(
  n_participants = c(400, 500, 600, 700, 800),
  mean_sessions  = c(3.0, 4.0, 5.0, 6.0, 8.0)
)

n_conditions <- nrow(param_grid)
n_cores      <- max(1L, detectCores() - 1L)   # leave one core for the OS

cat(sprintf(
  "Hardware: %d logical cores detected, using %d\n",
  detectCores(), n_cores
))
cat(sprintf(
  "Sweep: %d conditions x 500 simulations — progress bar in console\n\n",
  n_conditions
))

t_start <- proc.time()

# pbmclapply = mclapply + console progress bar.
# Each list element is one row of param_grid; mc.set.seed propagates the
# L'Ecuyer seed stream into each child process automatically.
results_list <- pbmcapply::pbmclapply(
  seq_len(n_conditions),
  function(k) {
    tryCatch(
      run_power_sim(
        n_participants = param_grid$n_participants[k],
        mean_sessions  = param_grid$mean_sessions[k],
        n_sims         = 500
      ),
      error = function(e) tibble(
        n_participants = param_grid$n_participants[k],
        mean_sessions  = param_grid$mean_sessions[k],
        n_sims_valid   = 0L,
        power          = NA_real_,
        power_lower    = NA_real_,
        power_upper    = NA_real_
      )
    )
  },
  mc.cores    = n_cores,
  mc.set.seed = TRUE
)

elapsed <- (proc.time() - t_start)[["elapsed"]]
cat(sprintf("\nDone. Total time: %s\n", format_time(elapsed)))

# mclapply returns try-error objects for crashed workers rather than raising;
# report them and drop before binding.
is_error <- sapply(results_list, inherits, "try-error")
if (any(is_error)) {
  message(sprintf(
    "%d worker(s) failed and will be excluded: conditions %s",
    sum(is_error),
    paste(which(is_error), collapse = ", ")
  ))
  results_list <- results_list[!is_error]
}

power_results <- bind_rows(results_list)
```

---

## 5. Results

### Power table

```{r results-table, echo=FALSE}
power_results |>
  mutate(
    power       = round(power,       3),
    power_lower = round(power_lower, 3),
    power_upper = round(power_upper, 3),
    meets_target = ifelse(power >= 0.80, "✓", "")
  ) |>
  arrange(n_participants, mean_sessions) |>
  rename(
    "N participants"    = n_participants,
    "Mean sessions"     = mean_sessions,
    "Valid sims"        = n_sims_valid,
    "Power"             = power,
    "Lower 95% CI"      = power_lower,
    "Upper 95% CI"      = power_upper,
    "≥ 80%?"            = meets_target
  ) |>
  kable(align = "rrrrrrr") |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed")) |>
  row_spec(
    which(power_results$power >= 0.80),
    background = "#d4edda"
  )
```

### Power curve

```{r power-curve, echo=FALSE}
ggplot(
  power_results,
  aes(
    x      = n_participants,
    y      = power,
    colour = factor(mean_sessions),
    fill   = factor(mean_sessions),
    group  = factor(mean_sessions)
  )
) +
  geom_ribbon(
    aes(ymin = power_lower, ymax = power_upper),
    alpha = 0.12, colour = NA
  ) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.5) +
  geom_hline(
    yintercept = 0.80,
    linetype   = "dashed",
    colour     = "grey35",
    linewidth  = 0.6
  ) +
  annotate(
    "text", x = 410, y = 0.815, label = "80% target",
    hjust = 0, size = 3.2, colour = "grey35"
  ) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, 1)) +
  scale_colour_brewer(
    palette = "Set2",
    name    = "Mean sessions\nper participant"
  ) +
  scale_fill_brewer(
    palette = "Set2",
    name    = "Mean sessions\nper participant"
  ) +
  labs(
    title    = "Power simulation: H1 — vibe_coding_pct × months interaction",
    subtitle = paste0(
      "MESI: 5pp accuracy decline per 6 months at 100% vs 0% vibe coding  |  ",
      "500 simulations per condition\n",
      "lme4 Wald t-test approximation of brms primary model"
    ),
    x = "Number of participants",
    y = "Power"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position  = "right",
    plot.subtitle    = element_text(size = 9, colour = "grey40"),
    panel.grid.minor = element_blank()
  )
```

### Recommendation

```{r recommendation, echo=FALSE}
# Find the smallest N that hits 80% at each session level
recommendation <- power_results |>
  filter(power >= 0.80) |>
  group_by(mean_sessions) |>
  slice_min(n_participants, n = 1) |>
  ungroup() |>
  arrange(mean_sessions)

if (nrow(recommendation) == 0) {
  cat("No tested condition reaches 80% power. Consider increasing N or sessions.")
} else {
  recommendation |>
    mutate(power = scales::percent(power, accuracy = 0.1)) |>
    select(
      "Mean sessions per participant" = mean_sessions,
      "Minimum N for ≥80% power"     = n_participants,
      "Achieved power"                = power
    ) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
}
```

---

## 6. Caveats

1. **Model approximation.** `lme4` Wald t-tests are a fast proxy for the `brms` Bayesian model (research_plan.md §5.2). Bayesian posteriors with weakly informative priors tend to be slightly wider, so true power under `brms` may be marginally lower. Validate a subset with the snippet below.

2. **Zero-correlation random effects (`||`).** The simulation uses `(1 + time_unit || participant_id)` rather than the correlated form `(1 + time_unit | participant_id)` specified in the primary brms model. With only 2 sessions per participant the correlation parameter is unidentifiable, causing widespread singular fits. The `||` form drops that correlation, removing one variance parameter and eliminating the warnings. The fixed-effect interaction t-value — the only quantity we care about here — is not materially affected by this simplification.

3. **Parallel RNG.** `RNGkind("L'Ecuyer-CMRG")` with `mc.set.seed = TRUE` gives each forked process its own independent random stream derived from the master seed, ensuring the results are fully reproducible.

4. **Session distribution.** Minimum 2 sessions, expected mode ~6 (roughly one session every 1–2 months over 12 months). The grid spans mean_sessions 3–8 to bracket pessimistic dropout through high engagement. Use the `mean_sessions = 3.0` row as the conservative bound.

5. **Vibe-coding distribution.** Simulated as bimodal (low: 0–25%, high: 50–100%). Less variance in the predictor → lower power.

6. **Tier 3–5 sparsity.** The simulation correctly reflects that most attempts are Tier 1–2. H3 (tier moderation) is **not** powered here — it is flagged as potentially exploratory in `research_plan.md §15`.

7. **Challenge random effects.** Simplified as iid noise. The full brms model uses `(1 | challenge_id)`; this minor simplification slightly inflates power estimates.

---

## 7. brms Validation Snippet

Run this on a single simulated dataset to confirm the lme4 approximation is reasonable. **Slow — not cached.**

```{r brms-validation, eval=FALSE}
library(brms)

dat_val <- simulate_dataset(n_participants = 200, mean_sessions = 2.5)

m_brms <- brm(
  accuracy_logit ~ vibe_coding_pct_c * time_unit +
    difficulty_tier_c +
    (1 + time_unit | participant_id),
  data   = dat_val,
  family = gaussian(),
  prior  = c(
    prior(normal(0, 1),   class = Intercept),
    prior(normal(0, 0.5), class = b),
    prior(normal(0, 0.5), class = sd),   # truncated to >0 automatically by brms
    prior(lkj(2),         class = cor)
  ),
  chains = 4, cores = 4, iter = 2000, warmup = 1000,
  seed   = 20260219
)

# Does the 95% CI for the interaction exclude zero?
fixef(m_brms)["vibe_coding_pct_c:time_unit", c("Estimate", "Q2.5", "Q97.5")]
```

---

## Session Info

```{r session-info, echo=FALSE}
sessionInfo()
```
